---
title: "IODS-Final"
author: "Ville Aro, ville.t.aro@helsinki.fi"
date: March 1, 2017
output:
  html_document:
    theme: united
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6
    fig_height: 4
    code_folding: hide
---

```{r, include=FALSE}
library(ggplot2)
library(GGally)
library(MASS)
library(corrplot)
library(tidyverse)
library(plotly)
library(dplyr)
library(FactoMineR)
require(ggplot2)

```

# Introduction and a brief look at my data

I am using a dataset from [UCI Machine Learning Repository](https://archive.ics.uci.edu/ml/datasets/STUDENT+ALCOHOL+CONSUMPTION). The dataset used in my analysis has combined information about the alcohol consumption of students of both maths and Portugese. Here are the variables that I chose and also an explanation how to measure the variable.

Variable | Description of variable and how it was measured
--------- | -----------------------------------------
G1 | first period grade (numeric: from 0 to 20) 
G2 | second period grade (numeric: from 0 to 20) 
G3 | final grade (numeric: from 0 to 20) 


```{r}
alc <- read.table("~/GitHub/IODS-Final/alcdata1", header=T, sep=",")

alcdata_ <- select(alc, -X)

dim(alcdata_)

colnames(alcdata_)

```


# Graphical overview

These are the variables that I decided to keep. Next I drew a multiplot of them where we can see better how they look.

```{r, include= FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

```

```{r}
g1 <- ggplot(alcdata_, aes(G1)) + geom_bar()+ theme_grey() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
g2 <- ggplot(alcdata_, aes(G2)) + geom_bar()+ theme_grey() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))
g3 <- ggplot(alcdata_, aes(G3)) + geom_bar()+ theme_grey() + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10))



multiplot(g1, g2, g3, cols=3)
```



```{r}
summary(alcdata_)

```


# My hypothesis

# The clustering model

# Results






